---
title: "Introduction to plotlycsv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to plotlycsv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`plotlycsv` is a lightweight R package designed to add a "Download CSV" button to Plotly modebars. This is particularly useful for static HTML documents (like those generated by Quarto or R Markdown) where you want to allow users to download the data behind a chart without requiring a Shiny server.

## Installation

You can install the development version of `plotlycsv` from GitHub (or local source):

```r
# If using pak:
# pak::pkg_install("RussellPolitzky/plotlycsv")

# If using devtools:
# devtools::install_github("RussellPolitzky/plotlycsv")
```

## Basic Usage

The primary function in the package is `add_plotly_export_csv()`. It acts as a decorator for any `plotly` htmlwidget.

### Using with ggplotly

Most users will use this in conjunction with `ggplotly()` to add export capabilities to their `ggplot2` charts.

```{r setup}
library(ggplot2)
library(plotly)
library(plotlycsv)
```

```{r example}
# Create a standard ggplot
p <- ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) + 
  geom_point() +
  labs(title = "Car Data: Weight vs MPG")

# Convert to plotly and add the export button
p_plotly <- ggplotly(p) %>% 
  add_plotly_export_csv(filename = "mtcars_data.csv")

# In a Quarto/RMarkdown doc, printing the object displays the chart with the button
# p_plotly
```

## Customizing the Export

`add_plotly_export_csv()` provides several options to control what data is exported:

- `filename`: The default filename suggested to the browser (defaults to `"plotly-data.csv"`).
- `visible_only`: If `TRUE` (the default), only traces currently visible in the chart (respecting legend toggles) will be exported.
- `include_text`: If `TRUE`, includes the hover text as a column in the CSV.
- `include_z`: Set to `TRUE` for 3D or heatmap charts if you want to include the `z` dimension.
- `include_customdata`: If your plotly object has `customdata` attached, set this to `TRUE` to include it.

### Example: Custom Filename and Hidden Traces

```r
ggplotly(p) %>% 
  add_plotly_export_csv(
    filename = "custom_export.csv",
    visible_only = FALSE # Export all data, even if hidden in the legend
  )
```

## How it Works

The package injects a small JavaScript payload into the Plotly widget using `htmlwidgets::onRender()`. 
When the "Download CSV" button in the modebar is clicked, the script:
1. Iterates through the currently rendered traces (`gd.data`).
2. Extracts the `x`, `y`, (and optionally `z`, `text`, `customdata`) values.
3. Generates a CSV string in the browser.
4. Triggers a client-side download using a Blob and a temporary link element.

Because the processing happens entirely in the browser, no connection to an R process or Shiny server is required after the HTML page is rendered.
